Chemical potential of liquids and mixtures via Adaptive Resolution Simulation

### 2014-05-11 writeDown/pub.jcp/chem-pot-jcp.tex
my revise

\item divide the space in three regions, one with atomistic resolution (AT) and one with coarse-grained (spherical) resolution (CG) interfaced by a smaller region with an hybrid treatment, which is usually called transition region or hybrid region.
\item couple the molecules in the different regions through a spatial interpolation formula on the forces:

@@

  $p_{\AT}+\rho_{0}\int_{\Delta}{\vect F}_{th}({\vect r})d{\vect r}=p_{\CG}$,
  where $p_{AT}$  is the target pressure of the atomistic system (region), $p_{CG}$ is the pressure of the coarse-grained model, $\rho_{0}$ is the target molecular density of the atomistic system (region)~\cite{prl12}. An additional locally acting thermostat is added to take care of the loss/gain of energy in the transition region.

@@

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{While the determination of $\omega_{\dof}$ is not required for the our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\thermo}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the Appendix)}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{AT}$ and $U_{CG}$ are the atomistic and coarse-grained potential.
\recheck{It must be noticed that the auxiliary Hamiltonian approach shall not be considered a Hamiltonian approach to adaptive resolution simulation. In fact, as discussed in Ref.\cite{prx} the equilibrium is imposed artificially and {\it per se} does not have any physical meaning (for more details, see discussion in the Appendix)}. 
  \recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.

The result above implies that $\omega_{\thermo}$ can be calculated in a straightforward way during the initial equilibration within in the standard GC-AdResS code. 
\recheck{It must be noticed that, within the AdResS scheme, an approach similar to the auxiliary Hamiltonian  has been recently proposed and applied to liquids ad mixtures (of toy models so far) by Potestio {\it et al.} \cite{h-adress-0, h-adress} (see also \cite{luigientropy} where such an approach is commented)}. At this point according to \eqref{mu}, if one knows $\mu_{\CG}$, then  GC-AdResS can automatically provide $\mu_{\AT}$. However we need to do one step more, in fact the quantity of interest is not the total chemical potential, but the excess chemical potential $\mu^{\exc}_{\AT}$ which corresponds to the expression of \eqref{mu} where the kinetic (ideal gas) part is subtracted. Regarding the kinetic part, one can notice that the contribution coming from the COM is the same for the coarse-grained and for the atomistic molecules, thus it is automatically removed in the calculation of \eqref{mu}.
The kinetic part of $\mu_{\AT}$ due to the rotational and vibrational degrees of freedom corresponds in our case to $\omega_{\dof}$ and \recheck{in principle can be calculated by hand (chemical potential of an isolated molecules). However such a calculation may become rather tedious for large and/or complex molecules but in our case it is actually not required.} In fact the \recheck{Gromacs implementation of AdResS} considers the removed degrees of freedom as phantom variables but thermally equilibrate them anyway \cite{simon-ch}. Thus the heat provided by the thermostat for the rotational and vibrational part is the same in the atomistic and coarse-graining molecules and is automatically removed in the difference. Finally, the calculation of $\mu^{\exc}_{\CG}$ can be done with standard methods, TI or IPM, which for simple spherical molecules, like those of the coarse-grained system, requires a negligible computational cost.
In conclusion, we have the final expression:
\begin{equation}
  \mu^{\exc}_{\AT}=
  \mu^{\exc}_{\CG}-
  \int_{\Delta}{F}_{th}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}
\label{finalmu}
\end{equation}

\section{Analytic Derivation of the $w_{\ext}$}\label{sec:tmp4}
\recheck{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We define a potential coupling the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where index $\alpha$ and $\beta$ denotes the atom position of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

@@

resolution~\eqref{eqn:peq} does not hold anymore and should be derived again.

@@

region.  The volume of the transition region is kept constant as if an ideal ``piston'' moves toward the CG region by an amount $\Delta L$.

@@

It must be noticed that also the displacements of the molecules are infinitesimal, so that it can be reasonably assumed that

@@

The notation $\langle\cdot\rangle$ in Eq.~\eqref{eqn:apptmp3} denotes the ensemble average.

@@

In Ref.~\cite{prx} we proved that under proper assumptions,
when the flat density profile is enforced by the thermodynamic force,
the chemical potential difference between the different resolutions
is given by
\begin{align}\label{eqn:mueq-d}
 \mu_\CG - \mu_\AT =   \omega_\thf + \omega_\dof + \omega_\ext
\end{align}
The same argument can be applied to the auxiliary Hamiltonian approach,
and yields  the chemical potential difference between the AT and CG resolutions
\begin{align}\label{eqn:mueq-h}
 \mu_\CG - \mu_\AT =   \omega^\hadress_\thf + \omega_\dof 
\end{align}
In the auxiliary Hamiltonian, we do not have the term $w_\ext$ in the above
formula (being this latter, in GC-AdResS, generated by the non conservative effect of the force interpolation).
By comparing~\eqref{eqn:mueq-d} with~\eqref{eqn:mueq-h}, we have the
relation
\begin{align}\label{eqn:hd-rel-2}
  \omega_\ext = \omega^\hadress_\thf - \omega_\thf,
\end{align}
which also relates the thermodynamic force of the auxiliary Hamiltonian AdResS
and GC-AdResS.\\

From Eq.~\eqref{eqn:hd-rel} and \eqref{eqn:hd-rel-2}, we
find the extra work of the thermostat in GC-AdResS being identical to the work of changing
representation of the Hamiltonian auxiliary approach:
\begin{align}\label{eqn:wextra}
  \omega_\ext = \omega_\res,
\end{align}
which basically proves the statement at the beginning of this section.  The ensemble
average on the R.H.S.~of Eq.~\eqref{eqn:w-rep} is performed in the
ensemble of the system treated with the potential interpolation approach, and the question is if the ensemble average is equivalent if it is performed in the simulation where the force
interpolation approach is used.  It is obvious that the spatial probability distribution corresponding to the system treated with the potential interpolation is consistent with the force
interpolation at least up to the 1st order as defined in Eq.\ref{order}.  It is also
possible to systematically obtain equivalence in the ensemble average operation  at higher orders of accuracy of the probability distribution, as, for example, it is done for the radial distribution function in Ref.~\cite{jctchan}.
However, here we do not consider higher order corrections, because it
has been numerically shown that actually the ensemble average of
$\vect F_\res$ dose not depend on in which ensemble it is
calculated~\cite{prx}. Therefore, we use Eq.~\eqref{eqn:wextra} to
calculate $\omega_\ext$, and measure the ensemble average by the
standard AdResS.  As previously reported, in the computer implementation, the CG
molecules also keep the atomistic degrees of freedom even though they
are in the CG region, therefore, the kinetic part of $\mu_\AT$ and
$\mu_\CG$ are identical, and $\omega_\dof$ vanishes. Therefore,
by inserting Eq.~\eqref{eqn:wextra} into \eqref{eqn:mueq-d}, we have
\begin{align}\label{eqn:tmp24}
  \mu_\AT^\exc = \mu_\CG^\exc - \int_\HY d\vect r \vect F_\thf(\vect r)
-\int_\HY d \vect r\,\nabla_{\vect r}w(\vect r)\,
\big\langle w(\vect r')\, [\sum_{\alpha,\beta}U^\AT(\vect r_\alpha-\vect r'_\beta) - U^\CG(\vect r-\vect r')]\big\rangle_{\vect r'; \vect r}
% -\int_\HY dr \nabla\omega(r)\,\big\langle \omega(r')\, [U^\AT(r-r') - U^\CG(r-r')]\big\rangle_{r'}
\end{align}
The extension of Eq.\ref{eqn:tmp24} to multicomponent systems is reported in the Appendix, while in the next section we apply the method to the calculation of $\mu^{ex}$ to liquids and mixtures.}

\section{Results and Discussion}
We have calculated $\mu^{ex}$ for different liquids and mixtures, choosing cases which are representative of a large class of systems. Hydrophobic solvation in methane/water and in ethane/water mixtures, hydrophilic solvation in urea/water, a balance of both in water/tert-Butyl alcohol (TBA) mixture, other liquids, e.g. pure methanol and DMSO (and their mixtures with water), non aqueous mixtures in TBA/DMSO and alkane liquids such as methane, ethane and propane. Moreover, systems as water/urea are commonly used as cosolvent of biological molecules \cite{nico-debashish} while systems as tert-Butyl alcohol/water play a key role in modern technology \cite{irata}, thus they are of high interest {\it per se}. All technical details of each simulation are presented in the Appendix.
Results are reported in Table \ref{table} where the comparison with values obtained using full atomistic TI and available experiments, at the same concentrations, of our calculation is made; in our previous work we have already shown that value of the chemical potential of liquid water obtained with IPM is well reproduced by GC-AdResS, however the computational cost of IMP was very large, thus we do not consider calculations done with IPM in this paper. 
The agreement with full atomistic TI simulations is satisfactory in all cases, and thus it proves the solidity of GC-AdResS in describing the essential thermodynamics of a large class of systems.
We also compare the obtained values with those available in literature \cite{vang,nico}. Although the concentration of the minor component in the mixtures that we consider, is higher than the concentrations considered in Refs.\cite{vang,nico}, we are anyway in the very dilute regime and thus the chemical potential should not change in a significant way; we have verified  such a supposed consistency. 
The chemical potential of $k$-th liquid's component in a mixture is calculated as \recheck{(see Appendix)}:
\begin{equation}
  \mu^{\exc,k}_{\AT}=
  \mu^{\exc,k}_{\CG}-
  \int_{\Delta}{\vect F}^k_{\thf}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r, k} d{\vect r}
% \mu^{ex,i}_{AT}=\mu^{ex,i}_{CG}-\int_{\Delta}{F}^{i}_{th}({\vect r})d{\vect r}-\int_{\Delta}\langle w\nabla_{x_{i}} w(x)(V_{AT}-V_{CG})\rangle_{\vect r}d{\vect r}
\label{eqn:tmp25}
\end{equation}
where ${F}^{k}_{\thf}(x)$ is the thermodynamic force applied to the molecules of the $k$-th component; this assures that, at the given concentration, the density of molecules of species $k$, in the transition region, is equivalent to the density of the same liquid's component in a reference full atomistic simulation.
\recheck{The ensemble average is taken over the position of the second molecule, provided that the first molecule is of species $k$, and located at position $\vect r$}.
% $\nabla_{x_{i}} w(x)$ is the gradient along $x$ (in a rectangular box) applied to the switching function $w(x)$ in the COM of molecules of component $i$.

@@

\appendix
\section{Technical details of the simulations}
The potential energy function and the force field parameters for all the molecules
were taken from GROMOS53A6 parameter set. Liquid water was described by the SPC
model~\cite{spc}, methanol was described by the model developed by Walser et al~\cite{walser},
urea by the model described in~\cite{urea}, tert-butyl alcohol by the parameter set of~\cite{tba}
and DMSO was described by the model given by Geerke et.al.~\cite{dmso1}. For liquid methanol simulations, 
GROMOS43A1 parameter set was used, as it was shown to be more accurate for calculating excess free energy of 
solvation of methanol in methanol~\cite{vang}.

@@

$\mu^{excess}$ is derived \cite{mu}. 

@@

 \mu_{iB} - \mu_{iA} = \int_{0}^{1} \left \langle \frac{\partial H}{\partial \lambda} \right \rangle_{\lambda} d\lambda 

@@

where $U_{i}$ is the interaction energy of particle $i$ with the remaining particles and $\langle . \rangle$

@@

where $\Delta G_{ele}$ is the free energy change associated with introducing the van der Waals interactions and 
$\Delta G_{vdw}$ is the free energy change associated with introducing Coulomb interactions. 

@@

\section{Technical Aspects of the Auxiliary Hamiltonian AdResS}
\recheck{In principle when the auxiliary Hamiltonian approach is used,
  one can perform microcanonical simulations and thus can avoid the
  use of a thermostat. In this case the thermodynamic force of the
  auxiliary Hamiltonian would not carry any effect of the thermostat and thus the difference between the work of the thermodynamic force of GC-AdResS and that of the auxiliary Hamiltonian is
  exactly the work that the thermostat does in GC-AdResS in order to compensate energy dissipation. The question is whether
  the energy is conserved in the auxiliary Hamiltonian approach. We
  have checked that the conservation holds for systems without
  electrostatics (methane,ethane,propane), thus for such systems the
  procedure is straightforward. Instead, for systems with
  electrostatic interactions, even for full atomistic simulations, due
  to the fact that the force fields are designed for employing the
  reaction field method, the energy cannot be conserved and the
  coupling to a thermostat is required. This is a well known problem
  reported in the manual of Gromacs. However, in our case, for both,
  the auxiliary Hamiltonian and GC-AdResS the energy drift due to the
  reaction field method is the essentially same because they have
  equivalent electrostatic interactions, thus the energy drift due to
  the use of the reaction field method is automatically removed when
  we consider the difference between the thermodynamic forces of the
  two approaches, that is the force of changing resolution.}

\section{Extension of the chemical potential derivation to multi-component systems}

@@

$\typeb$, provided that a molecule $\typea$ takes the position $\vect r$ (the same applies for other combinations on indices ${\vect r'}, {\vect r}, {\typea}, {\typeb}$.

>>>>

\item Divide the space in three regions, one with atomistic resolution (AT) and one with coarse-grained (spherical) resolution (CG) interfaced by a smaller region with an hybrid treatment, which is usually called transition region or hybrid region.
\item Couple the molecules in the different regions through a spatial interpolation formula on the forces:

@@

  $p_{\AT}+\rho_{0}\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}=p_{\CG}$,
  where $p_{\AT}$  is the target pressure of the atomistic system (region), $p_{\CG}$ is the pressure of the coarse-grained model, $\rho_{0}$ is the target molecular density of the atomistic system (region)~\cite{prl12}. An additional locally acting thermostat is added to take care of the loss/gain of energy in the transition region.

@@

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{While the determination of $\omega_{\dof}$ is not required for the our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\thermo}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in \bluec{the original} AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the \bluec{Appendix~\ref{app:tmp2}})}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{\AT}$ and $U_{\CG}$ are the atomistic and coarse-grained potential.
\recheck{It must be noticed that the auxiliary Hamiltonian approach shall not be considered a Hamiltonian approach to adaptive resolution simulation. In fact, as discussed in Ref.\cite{prx} the equilibrium is imposed artificially and {\it per se} does not have any physical meaning (for more details, see discussion in the \bluec{Appendix~\ref{app:tmp2}})}. 
  \recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.

The result above implies that $\omega_{\thermo}$ can be calculated in a straightforward way during the initial equilibration within in the standard GC-AdResS code. 
\recheck{It must be noticed that, within the AdResS scheme, an approach similar to the auxiliary Hamiltonian  has been recently proposed and applied to liquids and mixtures (of toy models so far) by Potestio {\it et al.} \cite{h-adress-0, h-adress} (see also \cite{luigientropy} where such an approach is commented)}. At this point according to \eqref{mu}, if one knows $\mu_{\CG}$, then  GC-AdResS can automatically provide $\mu_{\AT}$. However we need to do one step more, in fact the quantity of interest is not the total chemical potential, but the excess chemical potential $\mu^{\exc}_{\AT}$ which corresponds to the expression of \eqref{mu} where the kinetic (ideal gas) part is subtracted. Regarding the kinetic part, one can notice that the contribution coming from the COM is the same for the coarse-grained and for the atomistic molecules, thus it is automatically removed in the calculation of \eqref{mu}.
The kinetic part of $\mu_{\AT}$ due to the rotational and vibrational degrees of freedom corresponds in our case to $\omega_{\dof}$ and \recheck{in principle can be calculated by hand (chemical potential of an isolated molecules). However such a calculation may become rather tedious for large and/or complex molecules but in our case it is actually not required.} In fact the \recheck{Gromacs implementation of AdResS} considers the removed degrees of freedom as phantom variables but thermally equilibrate them anyway~\cite{simon-ch}. Thus the heat provided by the thermostat for the rotational and vibrational part is the same in the atomistic and coarse-graining molecules and is automatically removed in the difference. Finally, the calculation of $\mu^{\exc}_{\CG}$ can be done with standard methods, TI or IPM, which for simple spherical molecules, like those of the coarse-grained system, requires a negligible computational cost.
In conclusion, we have the final expression:
\begin{equation}
  \mu^{\exc}_{\AT}=
  \mu^{\exc}_{\CG}-
  \int_{\Delta}{F}_{th}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}
\label{finalmu}
\end{equation}

\section{Analytic Derivation of the $w_{\ext}$}\label{sec:tmp4}
\recheck{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We consider a potential coupling between the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where  $\alpha$ and $\beta$ denotes the atom indexes of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$ of molecule $i$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

@@

resolution~\eqref{eqn:peq} does not hold and should be derived again.

@@

region.  The volume of the transition region is kept constant as if it is an ideal ``piston'' that moves toward the CG region by an amount $\Delta L$.

@@

It must be noticed that also the displacements of the molecules are infinitesimal, so it can be reasonably assumed that

@@

The notation $\langle\cdot\rangle$ in Eq.~\eqref{eqn:apptmp3} denotes the ensemble average, which
will be specified soon.

@@

In Ref.~\cite{prx} we proved that under proper assumptions,
when the flat density profile is enforced by the thermodynamic force,
the chemical potential difference between the different resolutions
is given by
\begin{align}\label{eqn:mueq-d}
 \mu_\CG - \mu_\AT =   \omega_\thf + \omega_\dof + \omega_\ext
\end{align}
The same argument can be applied to the auxiliary Hamiltonian approach,
and yields  the chemical potential difference between the AT and CG resolutions
\begin{align}\label{eqn:mueq-h}
 \mu_\CG - \mu_\AT =   \omega^\hadress_\thf + \omega_\dof 
\end{align}
In the auxiliary Hamiltonian, we do not have the term $w_\ext$ in the above
formula \bluec{(being this latter, in GC-AdResS, generated by the non conservative effect of the force interpolation).}
By comparing~\eqref{eqn:mueq-d} with~\eqref{eqn:mueq-h}, we have the
relation
\begin{align}\label{eqn:hd-rel-2}
  \omega_\ext = \omega^\hadress_\thf - \omega_\thf,
\end{align}
which also relates the thermodynamic force of the auxiliary Hamiltonian AdResS
and GC-AdResS.\\

From Eq.~\eqref{eqn:hd-rel} and \eqref{eqn:hd-rel-2}, we
find the extra work of the thermostat in GC-AdResS being identical to the work of changing
representation of the auxiliary Hamiltonian approach:
\begin{align}\label{eqn:wextra}
  \omega_\ext = \omega_\res,
\end{align}
which basically proves the statement at the beginning of this section.  The ensemble
average on the R.H.S.~of Eq.~\eqref{eqn:w-rep} is performed in the
ensemble of the system treated with the potential interpolation approach, and the question is if the ensemble average is equivalent if it is performed in the simulation where the force
interpolation approach is used.  It is obvious that the spatial probability distribution corresponding to the system treated with the potential interpolation is consistent with the force
interpolation at least up to the first order.  It is also
possible to systematically obtain equivalence in the ensemble average operation  at higher orders of accuracy of the probability distribution, as, for example, it is done for the radial distribution function in Ref.~\cite{jctchan}.
However, here we do not consider higher order corrections, because it
has been numerically shown that actually the ensemble average of
$\vect F_\res$ dose not depend on in which ensemble it is
calculated~\cite{prx}. Therefore, we use Eq.~\eqref{eqn:wextra} to
calculate $\omega_\ext$, and measure the ensemble average by the
standard AdResS.  As previously discussed, in the Gromacs implementation, the CG
molecules also keep the atomistic degrees of freedom even though they
are in the CG region, therefore, the kinetic part of $\mu_\AT$ and
$\mu_\CG$ are identical, and $\omega_\dof$ vanishes. Therefore,
by inserting Eq.~\eqref{eqn:wextra} into \eqref{eqn:mueq-d}, we have
\begin{align}\label{eqn:tmp24}
  \mu_\AT^\exc = \mu_\CG^\exc - \int_\HY d\vect r \vect F_\thf(\vect r)
-\int_\HY d \vect r\,\nabla_{\vect r}w(\vect r)\,
\big\langle w(\vect r')\, [\sum_{\alpha,\beta}U^\AT(\vect r_\alpha-\vect r'_\beta) - U^\CG(\vect r-\vect r')]\big\rangle_{\vect r'; \vect r}
% -\int_\HY dr \nabla\omega(r)\,\big\langle \omega(r')\, [U^\AT(r-r') - U^\CG(r-r')]\big\rangle_{r'}
\end{align}
The extension of Eq.\ref{eqn:tmp24} to multicomponent systems is reported in the Appendix~\ref{app:tmp3}, while in the next section we apply the method to the calculation of $\mu^{ex}$ to liquids and mixtures.}

\section{Results and Discussion}
We have calculated $\mu^{ex}$ for different liquids and mixtures, choosing cases which are representative of a large class of systems. Hydrophobic solvation in methane/water and in ethane/water mixtures, hydrophilic solvation in urea/water, a balance of both in water/tert-Butyl alcohol (TBA) mixture, other liquids, e.g. pure methanol and DMSO (and their mixtures with water), non aqueous mixtures in TBA/DMSO and alkane liquids such as methane, ethane and propane. Moreover, systems as water/urea are commonly used as cosolvent of biological molecules \cite{nico-debashish} while systems as tert-Butyl alcohol/water play a key role in modern technology \cite{irata}, thus they are of high interest {\it per se}. All technical details of each simulation are presented in the Appendix~\ref{app:tmp1}.
Results are reported in Table \ref{table} where the comparison with values obtained using full atomistic TI and available experiments, at the same concentrations, of our calculation is made; in our previous work we have already shown that value of the chemical potential of liquid water obtained with IPM is well reproduced by GC-AdResS, however the computational cost of IMP was very large, thus we do not consider calculations done with IPM in this paper. 
The agreement with full atomistic TI simulations is satisfactory in all cases, and thus it proves the solidity of GC-AdResS in describing the essential thermodynamics of a large class of systems.
We also compare the obtained values with those available in literature \cite{vang,nico}. Although the concentration of the minor component in the mixtures that we consider, is higher than the concentrations considered in Refs.\cite{vang,nico}, we are anyway in the very dilute regime and thus the chemical potential should not change in a significant way; we have verified  such a supposed consistency. 
The chemical potential of $k$-th liquid's component in a mixture is calculated as \recheck{(see Appendix~\ref{app:tmp3})}:
\begin{equation}
  \mu^{\exc,k}_{\AT}=
  \mu^{\exc,k}_{\CG}-
  \int_{\Delta}{\vect F}^k_{\thf}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r, k} d{\vect r}
% \mu^{ex,i}_{AT}=\mu^{ex,i}_{CG}-\int_{\Delta}{F}^{i}_{th}({\vect r})d{\vect r}-\int_{\Delta}\langle w\nabla_{x_{i}} w(x)(V_{AT}-V_{CG})\rangle_{\vect r}d{\vect r}
\label{eqn:tmp25}
\end{equation}
where ${F}^{k}_{\thf}(x)$ is the thermodynamic force applied to the molecules of the $k$-th component; this assures that, at the given concentration, the density of molecules of species $k$, in the transition region, is equivalent to the density of the same liquid's component in a reference full atomistic simulation.
\recheck{The ensemble average is taken over the position of the second molecule, provided that the first molecule is of species $k$, and located at position $\vect r$}.
% $\nabla_{x_{i}} w(x)$ is the gradient along $x$ (in a rectangular box) applied to the switching function $w(x)$ in the COM of molecules of component $i$.

@@

\appendix
\section{Technical details of the simulations}\label{app:tmp1}
The potential energy function and the force field parameters for all the molecules
were taken from GROMOS53A6 parameter set. Liquid water was described by the SPC
model~\cite{spc}, methanol was described by the model developed by Walser et al~\cite{walser},
urea by the model described in~\cite{urea}, tert-butyl alcohol by the parameter set of~\cite{tba}
and DMSO was described by the model given by Geerke et.al.~\cite{dmso1}. For liquid methanol simulations, 
GROMOS43A1 parameter set was used, as it was shown to be more accurate for calculating excess free energy of 
solvation of methanol in methanol~\cite{vang}.

@@

$\mu^{\exc}$ is derived \cite{mu}. 

@@

 \mu^\exc_{iB} - \mu^\exc_{iA} = \int_{0}^{1} \left \langle \frac{\partial H}{\partial \lambda} \right \rangle_{\lambda} d\lambda 

@@

\bluec{where $U_{i}$ is the interaction energy of particle $i$ with the remaining particles} and $\langle \cdot \rangle$

@@

where $\Delta G_{vdw}$ is the free energy change associated with introducing the van der Waals interactions and 
$\Delta G_{ele}$ is the free energy change associated with introducing Coulomb interactions. 

@@

\section{Technical Aspects of the Auxiliary Hamiltonian AdResS}\label{app:tmp2}
\recheck{In principle when the auxiliary Hamiltonian approach is used,
  one can perform microcanonical simulations and thus can avoid the
  use of a thermostat. In this case, the thermodynamic force of the
  auxiliary Hamiltonian would not carry any effect of the thermostat, and thus the difference between the work of the thermodynamic force of GC-AdResS and that of the auxiliary Hamiltonian is
  exactly the work that the thermostat does in GC-AdResS in order to compensate energy dissipation. The question is whether
  the energy is conserved in the auxiliary Hamiltonian approach. We
  have checked that the conservation holds for systems without
  electrostatics (methane,ethane,propane), thus for such systems the
  procedure is straightforward. Instead, for systems with
  electrostatic interactions, even for full atomistic simulations, due
  to the fact that the force fields are designed for employing the
  reaction field method, the energy cannot be conserved and the
  coupling to a thermostat is required. This is a well known problem
  reported in the manual of Gromacs. However, in our case, for both,
  the auxiliary Hamiltonian and GC-AdResS the energy drift due to the
  reaction field method is the essentially same because they have
  equivalent electrostatic interactions, thus the energy drift due to
  the use of the reaction field method is automatically removed when
  we consider the difference between the thermodynamic forces of the
  two approaches, that is the force of changing resolution.}

\section{Extension of the chemical potential derivation to multi-component systems}\label{app:tmp3}

@@

$\typeb$, provided that a molecule $\typea$ takes the position $\vect r$ (the same applies for other combinations on indices ${\vect r'}, {\vect r}, {\typea}, {\typeb}$).
